---
title: "erika_model"
author: "Erika"
date: "2023-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this quarter's final lab, you will be using the knowledge of machine learning that you have gained this quarter to training models to predict dissolved inorganic carbon in water samples collected by the California Cooperative Oceanic Fisheries Investigations program (CalCOFI).

## Your Task:
- Acquire domain knowledge. (Dr. Satterthwaite's presentation)
- Explore the data.
- Preprocessing.
- Choose a model algorithm.
- Tune relevant parameters (Cross validation).
- Submit your prediction.

```{r include=FALSE}
library(caret) 
library(tidymodels) 
library(tidyverse)
library(dplyr) 
library(gbm)
library(xgboost)
library(ggpubr)
library(tictoc)
library(vip)
library(tictoc)
```


## Read in the data: 

This dataset was downloaded from the CalCOFI data portal. Bottle and cast data was downloaded and merged, then relevant variables were selected.

You will use the data contained in the train.csv file to train a model that will predict dissolved inorganic carbon (DIC) content in the water samples.

```{r}
dic_train <- read_csv("train.csv") %>% 
  select(-...13) %>% # remove column with all NAs
  rename(TA1 = TA1.x) %>% # rename column to match test data column name
  select(-id) # remove id column since it wouldn't add predictive value

dic_test <- read_csv("test.csv")

# don't need to split data since we already have a train and test set
```

Here's a short description of each variables: 

NO2uM - Micromoles Nitrite per liter of seawater
NO3uM - Micromoles Nitrate per liter of seawater
NH3uM - Micromoles Ammonia per liter of seawater
R_TEMP - Reported (Potential) Temperature in degrees Celsius
R_Depth - Reported Depth (from pressure) in meters
R_Sal - Reported Salinity (from Specific Volume Anomoly, M³/Kg)
R_DYNHT - Reported Dynamic Height in units of dynamic meters (work per unit mass)
R_Nuts - Reported Ammonium concentration
R_Oxy_micromol.Kg - Reported Oxygen micromoles/kilogram
PO4uM - Micromoles Phosphate per liter of seawater
SiO3uM - Micromoles Silicate per liter of seawater
TA1.x - Total Alkalinity micromoles per kilogram solution
Salinity1 - Salinity
Temperature_degC - Temp
DIC - Outcome

```{r}
# Preprocess the training data
dic_rec <- recipe(DIC ~ ., data = dic_train) %>%
  step_normalize(all_numeric(), -all_outcomes())

dic_rec_alt <- recipe(DIC ~ ., data = split_train) %>%
  step_normalize(all_numeric(), -all_outcomes())

# Set up cv folds
set.seed(123)
cv_folds <- dic_train %>% vfold_cv(v = 10)
```

# BOOSTED TREES

```{r}
# Create model specification
boost_spec_tune_1 <- boost_tree(
  trees = 3000, # ok to fix
  #tree_depth = tune(), min_n = tune(), loss_reduction = tune(), # tree params 
  #sample_size = tune(), mtry = tune(), # stochastic params 
  learn_rate = tune() # learning rate param
  ) %>% 
  set_engine("xgboost") %>%
  set_mode("regression") 


# Create tune grid
boost_grid_1 <- expand.grid(learn_rate = seq(0.0001, 0.3, length.out = 30))

# Create workflow
wf_boost_tune_1 <- workflow() %>%
  add_recipe(dic_rec) %>% # will automatically prep and bake
  add_model(boost_spec_tune_1) 

# Run the model
set.seed(123) # for reproducibility
doParallel::registerDoParallel()
tic("model run")
print("starting model tune 1...")
boost_rs_1 <- tune_grid(
wf_boost_tune_1,
DIC~., 
resamples = cv_folds,
grid = boost_grid_1,
metrics = metric_set(rmse)
)

print("...completing model tune 1")
toc()

# Collect metrics
boost_rs_df_1 <- boost_rs_1 %>%
  collect_metrics() %>%
  as.data.frame() %>%
  mutate(mean = as.numeric(mean)) %>%
  slice_min(mean, n = 5)
boost_rs_df_1
```
```{r}
# Create model specification
boost_spec_tune_2 <- boost_tree(
  trees = 3000, # ok to fix
  tree_depth = tune(), 
  min_n = tune(), 
  loss_reduction = tune(), # tree params 
  #sample_size = tune(), mtry = tune(), # stochastic params 
  learn_rate = 0.02078276 # learning rate param
  ) %>% 
  set_engine("xgboost") %>%
  set_mode("regression") 


# Grid specification
params_for_grid_2 <-
  dials::parameters(
    min_n(),
    tree_depth(),
    loss_reduction()
) 
# Create grid
set.seed(123) # for reproducibility 
boost_grid_2 <- dials::grid_max_entropy(params_for_grid_2,
                        size = 100 # test 100 combinations 
) 

# Create workflow
wf_boost_tune_2 <- workflow() %>%
  add_recipe(dic_rec) %>% # will automatically prep and bake
  add_model(boost_spec_tune_2) 

# Run the model
set.seed(123) # for reproducibility
doParallel::registerDoParallel()
tic("model run")
print("starting model tune 2...")
boost_rs_2 <- tune_grid(
wf_boost_tune_2,
DIC~., 
resamples = cv_folds,
grid = boost_grid_2,
metrics = metric_set(rmse)
)

print("...completing model tune 2")
toc()

#saveRDS(boost_rs_2, "boost_tree_specs.RDS")

# Collect metrics
boost_rs_df_2 <- boost_rs_2 %>%
  collect_metrics() %>%
  as.data.frame() %>%
  mutate(mean = as.numeric(mean)) %>%
  slice_min(mean, n = 5)
boost_rs_df_2

# 3 = min_n, 5 = tree_depth, 1.672370e-04 = loss_reduction
```

```{r}
# Create model specification
boost_spec_tune_3 <- boost_tree(
  trees = 3000, # ok to fix
  tree_depth = 5, 
  min_n = 3, 
  loss_reduction = 1.672370e-04, # tree params 
  sample_size = tune(), 
  mtry = tune(), # stochastic params 
  learn_rate = 0.02078276 # learning rate param
  ) %>% 
  set_engine("xgboost") %>%
  set_mode("regression") 


# Grid specification
params_for_grid_3 <-
  dials::parameters(
    sample_size = sample_prop(),
    finalize(mtry(), dic_train)
) 
# Create grid
set.seed(123) # for reproducibility 
boost_grid_3 <- dials::grid_max_entropy(params_for_grid_3,
                        size = 300 # test 100 combinations 
) 

# Create workflow
wf_boost_tune_3 <- workflow() %>%
  add_recipe(dic_rec) %>% # will automatically prep and bake
  add_model(boost_spec_tune_3) 

# Run the model
set.seed(123) # for reproducibility
doParallel::registerDoParallel()
tic("model run")
print("starting model tune 3...")
boost_rs_3 <- tune_grid(
wf_boost_tune_3,
DIC~., 
resamples = cv_folds,
grid = boost_grid_3,
metrics = metric_set(rmse)
)

print("...completing model tune 3")
toc()

#saveRDS(boost_rs_3, "boost_stoch_specs.RDS")

# Collect metrics
boost_rs_df_3 <- boost_rs_3 %>%
  collect_metrics() %>%
  as.data.frame() %>%
  mutate(mean = as.numeric(mean)) %>%
  slice_min(mean, n = 5)
boost_rs_df_3

# 5.426082, mtry 7, sample size 0.5081524
```

```{r}
# Create model specification
boost_spec_tune_4 <- boost_tree(
  trees = 3000, # ok to fix
  tree_depth = 5, 
  min_n = 3, 
  loss_reduction = 1.672370e-04, # tree params 
  sample_size = 0.5081524, 
  mtry = 7, # stochastic params 
  learn_rate = tune() # learning rate param
  ) %>% 
  set_engine("xgboost") %>%
  set_mode("regression") 


# Create tune grid
boost_grid_4 <- expand.grid(learn_rate = seq(0.0001, 0.3, length.out = 30))

# Create workflow
wf_boost_tune_4 <- workflow() %>%
  add_recipe(dic_rec) %>% # will automatically prep and bake
  add_model(boost_spec_tune_4) 

# Run the model
set.seed(123) # for reproducibility
doParallel::registerDoParallel()
tic("model run")
print("starting model tune 4...")
boost_rs_4 <- tune_grid(
wf_boost_tune_4,
DIC~., 
resamples = cv_folds,
grid = boost_grid_4,
metrics = metric_set(rmse)
)

print("...completing model tune 4")
toc()

#saveRDS(boost_rs_4, "boost_stoch_specs.RDS")

# Collect metrics
boost_rs_df_4 <- boost_rs_4 %>%
  collect_metrics() %>%
  as.data.frame() %>%
  mutate(mean = as.numeric(mean)) %>%
  slice_min(mean, n = 5)
boost_rs_df_4
```
```{r}
xgb_final <- finalize_model(boost_spec_tune_4, select_best(boost_rs_4))

set.seed(123)
# Make a final fit of the model with the data
train_fit <- xgb_final %>% fit(DIC ~ ., data = dic_train)
train_fit
```
```{r}
test_eval_fit <- predict(train_fit, new_data = dic_test)
test_eval_prob <- train_fit %>% 
  predict_numeric.model_fit(new_data = dic_test)

final_preds <- bind_cols(dic_test, test_eval_fit, test_eval_prob) %>% 
  rename(DIC = .pred) %>%
  select(id, DIC)

write.csv(final_preds, file = "submission.csv", row.names = FALSE)
```

