---
title: "erika_model"
author: "Erika"
date: "2023-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this quarter's final lab, you will be using the knowledge of machine learning that you have gained this quarter to training models to predict dissolved inorganic carbon in water samples collected by the California Cooperative Oceanic Fisheries Investigations program (CalCOFI).

## Your Task:
- Acquire domain knowledge. (Dr. Satterthwaite's presentation)
- Explore the data.
- Preprocessing.
- Choose a model algorithm.
- Tune relevant parameters (Cross validation).
- Submit your prediction.

```{r include=FALSE}
# Load packages
library(caret) 
library(tidymodels) 
library(tidyverse)
library(dplyr) 
library(gbm)
library(xgboost)
library(ggpubr)
library(tictoc)
library(vip)
```


## Read in the data: 

This dataset was downloaded from the CalCOFI data portal. Bottle and cast data was downloaded and merged, then relevant variables were selected.

You will use the data contained in the train.csv file to train a model that will predict dissolved inorganic carbon (DIC) content in the water samples.

#### RUN THE BELOW
```{r}
dic_train <- read_csv("train.csv") %>% 
  select(-...13) %>% # remove column with all NAs
  rename(TA1 = TA1.x) %>% # rename column to match test data column name
  select(-id) # remove id column since it wouldn't add predictive value

dic_test <- read_csv("test.csv")

# don't need to split data since we already have a train and test set
```

Here's a short description of each variables: 

NO2uM - Micromoles Nitrite per liter of seawater
NO3uM - Micromoles Nitrate per liter of seawater
NH3uM - Micromoles Ammonia per liter of seawater
R_TEMP - Reported (Potential) Temperature in degrees Celsius
R_Depth - Reported Depth (from pressure) in meters
R_Sal - Reported Salinity (from Specific Volume Anomoly, M³/Kg)
R_DYNHT - Reported Dynamic Height in units of dynamic meters (work per unit mass)
R_Nuts - Reported Ammonium concentration
R_Oxy_micromol.Kg - Reported Oxygen micromoles/kilogram
PO4uM - Micromoles Phosphate per liter of seawater
SiO3uM - Micromoles Silicate per liter of seawater
TA1.x - Total Alkalinity micromoles per kilogram solution
Salinity1 - Salinity
Temperature_degC - Temp
DIC - Outcome

```{r}
# Lets look at the data:
# Here are the variable correlations
corr <- Hmisc::rcorr(as.matrix(dic_train))
corr$r

# According to the correlations, it looks like most of the variables are highly correlated with DIC, so let's plot a few of the most correlated ones and compare them to the least correlated variables.
# First is how PO4uM (Phosphate concentration) is strongly positively correlated with DIC:
ggplot(dic_train, aes(x = PO4uM, y = DIC)) + 
  geom_point()

# This is how the reported oxygen concentrations affects DIC:
ggplot(dic_train, aes(x = R_Oxy_micromol.Kg, y = DIC)) + 
  geom_point()

# Latitude and longitude are relatively uncorrelated with DIC, so lets look at another uncorrelated variable: NO2 Concentrations
ggplot(dic_train, aes(x = NO2uM, y = DIC)) + 
  geom_point()

# Overall, it seems like it is safe to keep all the variables for training this model.
```


```{r}
# Preprocess the training data
dic_rec <- recipe(DIC ~ ., data = dic_train) %>%
  step_normalize(all_numeric(), -all_outcomes())

# Set up cv folds
set.seed(123) # for reproducibility
cv_folds <- dic_train %>% vfold_cv(v = 10)
```
#### RUN ALL THE ABOVE


# BOOSTED TREES
# CAN SKIP THE BELOW UNTIL INDICATED TO NOT SKIP - WAS USED FOR TUNING
```{r}
# # Create model specification for tuning the learn rate
# boost_spec_tune_1 <- boost_tree(
#   trees = 3000, # ok to fix
#   #tree_depth = tune(), min_n = tune(), loss_reduction = tune(), # tree params 
#   #sample_size = tune(), mtry = tune(), # stochastic params 
#   learn_rate = tune() # learning rate param
#   ) %>% 
#   set_engine("xgboost") %>%
#   set_mode("regression") 
# 
# 
# # Create tune grid
# boost_grid_1 <- expand.grid(learn_rate = seq(0.0001, 0.3, length.out = 30))
# 
# # Create workflow
# wf_boost_tune_1 <- workflow() %>%
#   add_recipe(dic_rec) %>% # will automatically prep and bake
#   add_model(boost_spec_tune_1) 
# 
# # Run the model
# set.seed(123) # for reproducibility
# doParallel::registerDoParallel()
# tic("model run")
# print("starting model tune 1...")
# boost_rs_1 <- tune_grid(
# wf_boost_tune_1,
# DIC~., 
# resamples = cv_folds,
# grid = boost_grid_1,
# metrics = metric_set(rmse)
# )
# 
# print("...completing model tune 1")
# toc()
# 
# # Collect metrics
# boost_rs_df_1 <- boost_rs_1 %>%
#   collect_metrics() %>%
#   as.data.frame() %>%
#   mutate(mean = as.numeric(mean)) %>%
#   slice_min(mean, n = 5)
# boost_rs_df_1

# # 0.02078276 = learn rate
```

```{r}
# # Create model specification for tuning the tree parameters, using best learn rate
# boost_spec_tune_2 <- boost_tree(
#   trees = 3000, # ok to fix
#   tree_depth = tune(), 
#   min_n = tune(), 
#   loss_reduction = tune(), # tree params 
#   #sample_size = tune(), mtry = tune(), # stochastic params 
#   learn_rate = 0.02078276 # learning rate param
#   ) %>% 
#   set_engine("xgboost") %>%
#   set_mode("regression") 
# 
# 
# # Grid specification
# params_for_grid_2 <-
#   dials::parameters(
#     min_n(),
#     tree_depth(),
#     loss_reduction()
# ) 
# # Create grid
# set.seed(123) # for reproducibility 
# boost_grid_2 <- dials::grid_max_entropy(params_for_grid_2,
#                         size = 100 # test 100 combinations 
# ) 
# 
# # Create workflow
# wf_boost_tune_2 <- workflow() %>%
#   add_recipe(dic_rec) %>% # will automatically prep and bake
#   add_model(boost_spec_tune_2) 
# 
# # Run the model
# set.seed(123) # for reproducibility
# doParallel::registerDoParallel()
# tic("model run")
# print("starting model tune 2...")
# boost_rs_2 <- tune_grid(
# wf_boost_tune_2,
# DIC~., 
# resamples = cv_folds,
# grid = boost_grid_2,
# metrics = metric_set(rmse)
# )
# 
# print("...completing model tune 2")
# toc()
# 
# # Collect metrics
# boost_rs_df_2 <- boost_rs_2 %>%
#   collect_metrics() %>%
#   as.data.frame() %>%
#   mutate(mean = as.numeric(mean)) %>%
#   slice_min(mean, n = 5)
# boost_rs_df_2
# 
# # 3 = min_n, 5 = tree_depth, 1.672370e-04 = loss_reduction
```

```{r}
# # Create model specification for tuning stochastic params, using found tree params and learn rate from other specs
# boost_spec_tune_3 <- boost_tree(
#   trees = 3000, # ok to fix
#   tree_depth = 5, 
#   min_n = 3, 
#   loss_reduction = 1.672370e-04, # tree params 
#   sample_size = tune(), 
#   mtry = tune(), # stochastic params 
#   learn_rate = 0.02078276 # learning rate param
#   ) %>% 
#   set_engine("xgboost") %>%
#   set_mode("regression") 
# 
# 
# # Grid specification
# params_for_grid_3 <-
#   dials::parameters(
#     sample_size = sample_prop(),
#     finalize(mtry(), dic_train)
# ) 
# # Create grid
# set.seed(123) # for reproducibility 
# boost_grid_3 <- dials::grid_max_entropy(params_for_grid_3,
#                         size = 300 # test 100 combinations 
# ) 
# 
# # Create workflow
# wf_boost_tune_3 <- workflow() %>%
#   add_recipe(dic_rec) %>% # will automatically prep and bake
#   add_model(boost_spec_tune_3) 
# 
# # Run the model
# set.seed(123) # for reproducibility
# doParallel::registerDoParallel()
# tic("model run")
# print("starting model tune 3...")
# boost_rs_3 <- tune_grid(
# wf_boost_tune_3,
# DIC~., 
# resamples = cv_folds,
# grid = boost_grid_3,
# metrics = metric_set(rmse)
# )
# 
# print("...completing model tune 3")
# toc()
# 
# # Collect metrics
# boost_rs_df_3 <- boost_rs_3 %>%
#   collect_metrics() %>%
#   as.data.frame() %>%
#   mutate(mean = as.numeric(mean)) %>%
#   slice_min(mean, n = 5)
# boost_rs_df_3
# 
# # 5.426082, mtry 7, sample size 0.5081524
```
# CAN SKIP THE ABOVE - WAS FOR TUNING

# BELOW THIS IS ABSOLUTELY NECESSARY TO RUN
```{r}
# Create model specification tuning learn rate one last time after tuning all other params, then select best from this
boost_spec_tune_4 <- boost_tree(
  trees = 3000, # ok to fix
  tree_depth = 5, 
  min_n = 3, 
  loss_reduction = 1.672370e-04, # tree params 
  sample_size = 0.5081524, 
  mtry = 7, # stochastic params 
  learn_rate = tune() # learning rate param
  ) %>% 
  set_engine("xgboost") %>%
  set_mode("regression") 


# Create tune grid
boost_grid_4 <- expand.grid(learn_rate = seq(0.0001, 0.3, length.out = 30))

# Create workflow
wf_boost_tune_4 <- workflow() %>%
  add_recipe(dic_rec) %>% # will automatically prep and bake
  add_model(boost_spec_tune_4) 

# Run the model
set.seed(123) # for reproducibility
doParallel::registerDoParallel()
tic("model run")
print("starting model tune 4...")
boost_rs_4 <- tune_grid(
wf_boost_tune_4,
DIC~., 
resamples = cv_folds,
grid = boost_grid_4,
metrics = metric_set(rmse)
)

print("...completing model tune 4")
toc()

# Collect metrics
boost_rs_df_4 <- boost_rs_4 %>%
  collect_metrics() %>%
  as.data.frame() %>%
  mutate(mean = as.numeric(mean)) %>%
  slice_min(mean, n = 5)
boost_rs_df_4
```
```{r}
# Select best model from retuning learn rate
xgb_final <- finalize_model(boost_spec_tune_4, select_best(boost_rs_4))

# Make a final fit of the model with the data
set.seed(123) # for reproducibility
train_fit <- xgb_final %>% fit(DIC ~ ., data = dic_train)
```
```{r}
# Use final training fit to predict on test data
test_eval_fit <- predict(train_fit, new_data = dic_test)

# Assemble predictions
final_preds <- bind_cols(dic_test, test_eval_fit) %>% 
  rename(DIC = .pred) %>%
  select(id, DIC)

#write.csv(final_preds, file = "submission.csv", row.names = FALSE) # create prediction file
```

