---
title: "jared_model"
author: "Jared Petry"
date: "2023-03-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this quarter's final lab, you will be using the knowledge of machine learning that you have gained this quarter to training models to predict dissolved inorganic carbon in water samples collected by the California Cooperative Oceanic Fisheries Investigations program (CalCOFI).

## Your Task:
- Acquire domain knowledge. (Dr. Satterthwaite's presentation)
- Explore the data.
- Preprocessing.
- Choose a model algorithm.
- Tune relevant parameters (Cross validation).
- Submit your prediction.

```{r include=FALSE}
library(caret) 
library(tidymodels) 
library(tidyverse)
library(dplyr) 
library(gbm)
library(xgboost)
library(ggpubr)
library(tictoc)
library(vip)
library(Hmisc)
```


## Read in the data: 

This dataset was downloaded from the CalCOFI data portal. Bottle and cast data was downloaded and merged, then relevant variables were selected.

You will use the data contained in the train.csv file to train a model that will predict dissolved inorganic carbon (DIC) content in the water samples.

```{r}
dic_train <- read_csv("train.csv") |> select(-...13) # take out blank column
dic_test <- read_csv("test.csv")
```

Here's a short description of each variables: 

NO2uM - Micromoles Nitrite per liter of seawater
NO3uM - Micromoles Nitrate per liter of seawater
NH3uM - Micromoles Ammonia per liter of seawater
R_TEMP - Reported (Potential) Temperature in degrees Celsius
R_Depth - Reported Depth (from pressure) in meters
R_Sal - Reported Salinity (from Specific Volume Anomoly, M³/Kg)
R_DYNHT - Reported Dynamic Height in units of dynamic meters (work per unit mass)
R_Nuts - Reported Ammonium concentration
R_Oxy_micromol.Kg - Reported Oxygen micromoles/kilogram
PO4uM - Micromoles Phosphate per liter of seawater
SiO3uM - Micromoles Silicate per liter of seawater
TA1.x - Total Alkalinity micromoles per kilogram solution
Salinity1 - Salinity
Temperature_degC - Temp
DIC - Outcome

**Heads up: one variable is differing in the train set: TA1.x in train set and TA1 in test set... Also column 13 is blank so I am removing it from the start

### Data exploration: 

```{r}
# let's see which variables are most correlated with variable DIC 

cor_w_dic <- function(var) {
  cor <- cor(dic_train$DIC, dic_train$var)
  print(paste("the correlation between", var, "and dissolved inorganic carbon is:", cor))
}

# test the function with one variable
cor_w_dic(Lat_Dec)

corr <- rcorr(as.matrix(dic_train))
corr$r

```
As we can see from this correlation matrix, the variables that are highly correlated with DIC are: 
- NO3uM
- R_TEMP
- R_Depth
- R_Sal
- R_DYNHT
- R_Oxy_micromol.Kg 
- PO4uM 
- SiO3uM 
- TA1.x 
- Salinity1

Variables that are NOT highly correlated with DIC: 
- id 
- Lat_Dec 
- Lon_Dec 
- NO2uM
- NH3uM
- R_Nuts

Create some cross validation folds for the training data: 
```{r}
folds <- vfold_cv(data = dic_train, 
                  v = 10,
                  strata = DIC)
```


## Pick a model: 

Here are the types of models that we have worked with in this class: 
linear regression 
- lab 1: used to predict the price of a pumpkin (linear_reg() function)
regularized multivariate polynomial regression 
- used the same as above with step_poly and poly_spec?? I am confused how different this would be or if this would be possible with many variables
classification and logistic regression 
- not applicable here because we are predicting a continuous variable
k nearest neighbors 
- could be a good option?

decision tree 
bagged forest
random forest 
boosted trees

We are not doing classification, we are trying to predict an actual number for DIC, so 

### Let's try a K-nearest neighbor model 

Create a recipe 
```{r}
dic_recipe <- recipe(DIC ~ ., data = dic_train)
# potentially add some pre-processing steps??
```





